The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-14 talk/vmil2012/paper.tex
Improve related work, evaluation and conclusion sections

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{find better names for JIT front- and backend}
\section{Introduction}

@@

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need to
store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

@@

\section{Evaluation}
\label{sec:evaluation}
\todo{improve the table formatting}
\todo{give a reference to the benchmark scripts to make things repeatable}

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the machine code
backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

From the mentioned benchmarks we collected different datasets to evaluate the
frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section. We want to point out three
aspects of guards in particular
\begin{itemize}
  \item Guards are very common operations in traces.
  \item There is overhead associated with guards.
  \item Guard failures are local and rare.
\end{itemize}

@@

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations which is between 69.4\% and 83.89\% of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar, as could be assumed based on
Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
most of the guards, but after the optimization pass guards still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

Why the efficient storing of the \texttt{resume data} is a central concern in the design
of guards is illustrated by Figure~\ref{fig:backend_data}. This figure shows
the size of the compressed \texttt{resume data}, the approximated size of
storing the \texttt{resume data} without compression and
an approximation of the best possible compression of the resume data by
compressing the data using the
\texttt{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently through using the techniques described above. On the other hand
comparing the results to the xz compression which only needs between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good compressed representation of the compressed resume data for the
large amount of guards present in the traces.

@@

After the guard is patched
failures execute the new bridge instead of jumping to the trampoline and returning to the interpreter. Hence the
numbers presented for guards that have a bridge represent the
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also of all failing guards a few fail extremely often
and most fail rarely. The results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

Guards as described are a concept associated with tracing just-in-time
compilers to represent possible divergent control flow paths.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used in there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions the plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of Fragment Linking.
Once a new hot trace is emitted into the fragment cache it is linked to side
exit that led to the compilation. Fragment Linking avoids the performance
penalty involved in leaving the compiled and it to remove the compensation
code used when restoring the machine state on a side exit.

Gal et. al~\cite{Gal:2006} describe that in the HotpathVM they experimented
with having one generic compensation code block, like the RPython JIT, that
uses a register variable mapping to restore the interpreter state. Later this
was replaced by generating compensation code for each guard which produced a
lower overhead in their benchmarks. HotpathVM also records secondary traces
starting from failing guards that are connected directly to the original trace.
Secondary traces are compiled by first restoring the register allocator state to
the state at the side exit. The information is retrieved from a mapping stored
in the guard that maps machine level registers and stack to Java level stack
and variables.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation typically found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
Based on the observation that guards are a frequent operation in traces and
that they do not fail often, we described how they have been implemented in the
high and low level components of RPython's tracing JIT compiler.

Finally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments confirmed that guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, but it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode compressed state representation.
The compressed state representation reduces the memory footprint of rarely
used data.

>>>>

%___________________________________________________________________________
\todo{find a better name for \texttt{low-level resume data}}
\todo{find better names for JIT front- and backend}
\todo{mention somewhere that it is to be expected that most guards do not fail}
\section{Introduction}

@@

Our aim is to help understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

@@

\section{Evaluation}
\label{sec:evaluation}
\todo{improve the table formatting}

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the machine code
backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
tools used to run and evaluate the benchmarks including the patches applied to
the PyPy sourcecode can be found in the repository for this
paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

From the mentioned benchmarks we collected different datasets to evaluate the
frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section. We want to point out three
aspects of guards in particular:
\begin{itemize}
  \item Guards are very common operations in traces.
  \item There is overhead associated with guards.
  \item Guard failures are local and rare.
\end{itemize}

@@

Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

Why the efficient storing of the \texttt{resume data} is a central concern in the design
of guards is illustrated by Figure~\ref{fig:resume_data_sizes}. This figure shows
the size of the compressed \texttt{resume data}, the approximated size of
storing the \texttt{resume data} without compression and
an approximation of the best possible compression of the resume data by
compressing the data using the
\texttt{xz} compression tool, which is a ``general-purpose data compression
software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently using the techniques described earlier. On the other hand
comparing the results to the xz compression which only needs between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good, compressed representation of the resume data for the
large amount of guards present in the traces.

@@

The numbers presented for guards that have a bridge represent the
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
and most fail rarely. The results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

Guards, as described, are a concept associated with tracing just-in-time
compilers to represent possible divergent control flow paths.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions the plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of Fragment Linking.
Once a new hot trace is emitted into the fragment cache it is linked to side
exit that led to the compilation of the fragment. Fragment Linking avoids the
performance penalty involved in leaving the compiled code. Fragment linking
also allows to remove compensation code associated to the linked fragments that
would have been required to restored the execution state on the side exit.

Gal et. al~\cite{Gal:2006} describe how in the HotpathVM they experimented
with having one generic compensation code block, like the RPython JIT, that
uses a register variable mapping to restore the interpreter state. Later this
was replaced by generating compensation code for each guard which produced a
lower overhead in their benchmarks. HotpathVM also records secondary traces
starting from failing guards that are connected directly to the original trace.
Secondary traces are compiled by first restoring the register allocator state to
the state at the side exit. The information is retrieved from a mapping stored
in the guard that maps machine level registers and stack to Java level stack
and variables.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
Based on the observation that guards are a frequent operation in traces and
that they do not fail often, we described how they have been implemented in the
high and low level components of RPython's tracing JIT compiler.

Additionally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments confirmed that guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, but it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode a compressed state representation.
The compressed state representation reduces the memory footprint of rarely
used data.
